Make any stupid plot you want

### 2021-01-22 draft/sections/intro.tex
finished out k
transition function
pulled more spivak back in

@@

\subsubsection{Data Type and Structure}
\label{sec:intro_data_structure}
\begin{figure}[h!]
 \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
\caption{Keys are unique lookup values used to find individual observations in the dataset. Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube. Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
\label{fig:munzner_datatypes}
\end{figure}

@@

>>>>

\subsection{what are we doing}
constraints: toplogy of dataset + structure on the variables
implementation: be fully feature
we can haz compose the things into larger structure. \& has framwework to do larger things w/ algebraic formalizations.

@@

\subsubsection{Data Type and Structure}
%% mention tidy/spivak first, set up that we use variable, record, key/index notation
\label{sec:intro_data_structure}
\begin{figure}[h!]
 \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
\caption{Keys are unique lookup values used to find individual observations in the dataset. Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube. Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
\label{fig:munzner_datatypes}
\end{figure}

@@

Throughout this form we will be using the following conventions to discuss data:

\begin{description}
    \item[variable]
    \item[index]  
    \item[record]
    \item[value]  
\end{description}

### 2021-01-22 draft/sections/math.tex
pulled more spivak back in

\subsection{Data Space}
\label{sec:data}
%%% 
We build on Butler's fiber bundle driven abstraction of data formats\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989},  ....

Fiber bundles are structures $(E,\,B,\,\pi ,\,F)$ where $E$, $F$, and $K$ are topological spaces that encode 
\begin{itemize}
\item the properties of the variables in the fiber $F$ (\ref{sec:data_fiber})
\item the continuity of the records in the base space $K$ (\ref{sec:data_base})
\item collections of records $\tau$ (\ref{sec:data_section}). 
\end{itemize}

where $E$ is the total space of data that $F$ lives in. The bundle is the projection map $\pi$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
    \end{tikzcd}
\end{equation}

that binds the variable space $F$ to the base space $K$.

%%probbaly need some more words here maybe, maybe pull that first picture back up now

The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified. In the trivial bundle $E=K\times F$ such that the total space $E$ falls out and discussion can be limited to the $K$ and $F$ spaces.

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.

%%push straight math generalization up
%% We define w/ mapping, as Spivak

%%push down to an example, knock it back to semi concrete
For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
\begin{equation}
    F = U_{temp}
\end{equation}
Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb{R}^+\}$ 
\begin{equation}
F =  U_{temp} \times U_{time}
\end{equation}
such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$. The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90^\circ \leq y \leq 90^\circ, -180^\circ \leq x -180^\circ\}$: %add comment about latitude and longitude
\begin{equation}
    F = U_{temp} \times U_{time} \times U_{loc}
\end{equation}
such that one record $r$ in the fiber would have the form $r=(temperature,\, time,\, (latitude,\, longitude))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
    F_2 &= (\textit{location},\, U_{loc},\, M_2)
\end{align}%the fiber is the set, variable is the tuple such that there is a map from the index i in F to the tuple (name, M_0)  for every U_{type} we can construct a map to the name and the type as described by Spivak 
 or more generally (name, type, monoid action)
 \begin{equation}
    F_i = (\textit{name},\, U_{type},\, M_{i})
 \end{equation}
where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

@@

\subsubsection{Case Study: GHCN}
\textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\

Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations. Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected

% break out the mbox/clean up the typesetting
The fiber bundle model is flexible enough to express some of the many different forms that temperature data can come in.

\begin{figure}[ht!]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
        %% add box around neighboring P and Map
        \label{fig:base_example_line}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
        \label{fig:base_example_plane}
    \end{subfigure}
    \label{fig:base_example}
    \caption{These two datasets have the same fiber of temperature but different base spaces. In figure~\ref{fig:base_example_line} the temperature values are 1D continuous, while in figure~\ref{fig:base_example_plane} the temperature values are 2D continuous.}
\end{figure}

The datasets in figure~\ref{fig:base_example} have identical fibers that encode a set of temperature values. In figure~\ref{fig:base_example_line} the temperatures lie on a line such that a section could return a timeseries or a distribution. In figure~\ref{fig:base_example_plane}, the temperatures lie on a 2D continuous plane; a section could return a map or contour. Because the fiber is 1D, it does not encode metadata

\begin{figure}[ht!]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
        \label{fig:fiber_example_plane}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
        \label{fig:fiber_example_cube}
    \end{subfigure}
    \label{fig:fiber_example}
    \caption{The fiber is expanded to include metadata fields that describe the semantics of $K$. In figure~\ref{fig:fiber_example_plane} the fiber is \textrm{temperature} $\times$ \textrm{time} and in figure~\ref{fig:fiber_example_cube} the fiber is \textrm{temperature}$\times$ \textrm{latitude} $\times$ {longitude}}.
\end{figure}

To encode the metadata, the fiber is expanded as illustrated in figure~\ref{fig:fiber_example}. The fiber in figure~\ref{fig:fiber_example_plane} is the cartesian product of the space of possible temperature values in degrees celsius and space of possible time values
\begin{equation}
F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
\label{eq:fiber_plane}
\end{equation}

while the fiber in figure~\ref{fig:fiber_example_cube} is the cartesian product of temperature, latitude, and longitude
\begin{equation}
F = [temp_{min}, temp_{max}] \times [-90, 90] \times [-180, 180]
\label{eg:fiber_cube}
\end{equation}

such that $E$ is the space of all possible points in $F$.

\begin{figure}[ht!]
    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
    \label{fig:fiber_example_section}
    \caption{The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points. $\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.  
\end{figure}
Given the fiber described in equation~\ref{eq:fiber_plane}, the sections $\tau_{1}$ and $\tau_{2}$ in figure~\ref{fig:fiber_example_section} return tuples of the form
\begin{equation}
\tau(k) = (k, (temperature, time))
\end{equation}
such that sections with the constraint that time is monotonic return a timeseries.

>>>>

\subsection{Data Space}
\label{sec:data}
We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough to encode all the types of data described in section~\ref{sec:intro_data_structure} and many more. We model data as the fiber bundle $(E,\,B,\,\pi ,\,F)$, where $E$, $F$, and $K$ are topological spaces that encode 
\begin{itemize}
\item the properties of the variables in the fiber $F$ (\ref{sec:data_fiber})
\item the continuity of the records in the base space $K$ (\ref{sec:data_base})
\item collections of records $\tau$ (\ref{sec:data_section}). 
\end{itemize}

and $E$ is the total space of data that $F$ lives in. The bundle is the projection map $\pi$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
    \end{tikzcd}
\end{equation}

that binds the variable space $F$ to the base space $K$. The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $E=K\times F$ such that extra structure in the total space $E$ falls out and discussion can be focused on the fiber and base space.

\begin{figure}[ht!]
    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
    \label{fig:data_base_space}
    \caption{}.  
\end{figure}
 We propose fiber bundles as an underlying data representation because they provide a way to decouple the components of a dataset. In figure~\ref{fig:}, we have two datasets $\tau_1$ and $\tau_2$ that have the same indexing structure $K$ and and the same schema $F$ and we can independently codify which properties of each component need to be preserved in the visualization.

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
We use Spivak's description of schemas in his categorical formulation of databases \cite{spivakSIMPLICIALDATABASES} as our fiber space. He describes the variables in a dataset as having names $C$ and datetypes $D$. Spivak then constructs a set $U_{\sigma}$ that is the universal set of values 
\begin{tikzcd}
    U_{\sigma} \arrow[r] \arrow[d, "\pi_{\sigma}"'] & U \arrow[d, "\pi"] \\
    C \arrow[r, "\sigma"']                          & \textbf{DT}       
\end{tikzcd}

restricted to values of types $\mathbf{DT}$ bound to variable names $C$. Therefore the fiber is for a one variable dataset is
\begin{equation}
    F = U_{\sigma}
\end{equation}
where $\sigma$ is the schema binding variable name to variable type. Therefore a dataset with multiple variables $c \in C$ has a fiber that is the cartesian cross product of $U_{\sigma}$ restricted over $c$ spaces:

\begin{equation}
F = U_{\sigma(c_1)}\times \ldots U{\sigma{c_i}} \ldots\timesU_{sigma(c_n)}
\end{equation}

which is equivalent to

\begin{equation}
    F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}

%%somehoe close out this piece.

For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
\begin{equation}
    F = U_{temp}
\end{equation}
Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb{R}^+\}$ 
\begin{equation}
F =  U_{temp} \times U_{time}
\end{equation}
such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$. The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90^\circ \leq y \leq 90^\circ, -180^\circ \leq x -180^\circ\}$: %add comment about latitude and longitude
\begin{equation}
    F = U_{temp} \times U_{time} \times U_{loc}
\end{equation}
such that one record $r$ in the fiber would have the form $r=(temperature,\, time,\, (latitude,\, longitude))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
    F_2 &= (\textit{location},\, U_{loc},\, M_2)
\end{align}

@@
